---
title: "FASTA Quick Stats"
author: "Emily Koehler"
output:
  html_document:
    df_print: paged
---

# FASTA Quick Stats


This notebook contains a few quick code chunks to get a general overview of a FASTA file, such as:



- Number of sequences (chromosomes, scaffolds, or contigs)
- Sequence lengths
- Total number of nucleotides
- GC content per sequence
- Number and percentage of ambiguous bases (`N`s)
- A peek at sequence names to see if they're labeled as chromosomes or scaffolds




Quick sanity check—this just confirms the file starts with a `>` header line (e.g., `>1`, `>chr1`) and is followed by base content (A, T, C, G, maybe some Ns).


```{r, message = FALSE, warning = FALSE}
# Load libraries
library(BiocManager)
library(Biostrings)

# File path
file_path <- ("/Users/emilykoehler/Downloads/128_Mmul_10.fasta")

#Preview file
readLines(file_path, n = 10)
```




Load FASTA into memory


```{r}
# Load the FASTA as DNAStringSet
fasta <- readDNAStringSet(filepath = file_path)

# Show basic info
fasta
```



Basic Sequence Stats


```{r}
# Number of sequences
num_seqs <- length(fasta)

# Sequence lengths
seq_lengths <- width(fasta)

# Total genome length
total_length <- sum(seq_lengths)

# Preview
cat("Number of sequences:", num_seqs, "\n")
cat("Total length (bp):", total_length, "\n")
```




GC Content per sequence


```{r}
# GC content per sequence
gc_content <- letterFrequency(fasta, letters = c("G", "C"), as.prob = TRUE)
gc_percent <- rowSums(gc_content) * 100

# Optional: round it
gc_percent <- round(gc_percent, 2)

# Add to summary table
summary_df <- data.frame(
  Sequence_ID = names(fasta),
  Length = seq_lengths,
  GC_Content_Percent = gc_percent,
  stringsAsFactors = FALSE
)

head(summary_df)
```




Ambiguous Bases (N Content)

```{r}
# Count N bases per sequence (ambiguous bases)
n_counts <- letterFrequency(fasta, letters = "N", as.prob = FALSE)
total_Ns <- sum(n_counts)
percent_N <- round(100 * total_Ns / total_length, 2)

cat("Total ambiguous (N) bases:", total_Ns, "\n")
cat("Percent of genome that is ambiguous (N):", percent_N, "%\n")
```


Sequence Name Overview

```{r}
# Preview just the first and last few sequence names
cat("First few sequence names:\n")
print(head(names(fasta), 5))

cat("\nLast few sequence names:\n")
print(tail(names(fasta), 5))
```















